Safe Transferable Regions

### 2016-07-07 paper/introduction.tex
Added scala link. Explained T@R advantages.

\paragraph{Performance.}
When the transferred data structures are large, as is the case with many streaming big-data
analysis systems, garbage collection overhead becomes
significant~\cite{Broom:HotOS}.  Furthermore, in a distributed
dataflow system, the GC pause at one node can have a cascading adverse
effect on the performance of other nodes~\cite{Broom:HotOS,harris15}:
a GC pause at an upstream actor can block downstream actors that are waiting for
messages.  However, often much of the GC overhead results from the
collector performing avoidable or unproductive work.  For the example
in Fig.~\ref{fig:motivating-eg}, GC might repeatedly traverse the
\C{map} data structure, although its objects cannot be collected until a
suitable timing message arrives.
There has been increasing interest in off-heap memory management for better performance
in the context of several  systems and languages (e.g., Spark~\cite{SPARK:2015}, Scala, Rust).
Several systems (e.g.,~\cite{TRILL:2014}) resort to using ``free object pools'' to partially alleviate this problem, in
the absence of language support.

>>>>

\paragraph{Performance.}
When the transferred data structures are large, as is the case with many streaming big-data
analysis systems, garbage collection overhead becomes
significant~\cite{Broom:HotOS}.  Furthermore, in a distributed
dataflow system, the GC pause at one node can have a cascading adverse
effect on the performance of other nodes~\cite{Broom:HotOS,harris15}:
a GC pause at an upstream actor can block downstream actors that are waiting for
messages.  However, often much of the GC overhead results from the
collector performing avoidable or unproductive work.  For the example
in Fig.~\ref{fig:motivating-eg}, GC might repeatedly traverse the
\C{map} data structure, although its objects cannot be collected until a
suitable timing message arrives.
There has been increasing interest in off-heap memory management for better performance
in the context of several  systems and languages (e.g.,
Spark~\cite{SPARK:2015}, Scala~\cite{ScalaOffHeap}, Rust).
Several systems (e.g.,~\cite{TRILL:2014}) resort to using ``free object pools'' to partially alleviate this problem, in
the absence of language support.

### 2016-07-07 paper/type-system.tex
Added scala link. Explained T@R advantages.

To construct objects of the \C{Pair} class, its type and region
parameters need to be instantiated with core types ($T$) and concrete
region identifiers\footnote{Region variables ($\rho$) and region
identifiers ($\rgn$) belong to the same syntactic class. Region
identifiers are to region variables, as types ($T$) are to type
variables ($a$)} ($\rgn$), respectively. For example:
\begin{codejava}
letregion $\rgn_0$ in
  let snd = new Object<$\rgn_0$>() in
  letregion $\rgn_1$ in
    let fst = new Object<$\rgn_1$>() in
    let p = new Pair<Object,Object><$\rgn_1$,$\rgn_1$,$\rgn_0$>
                  (fst,snd);
\end{codejava}
%\vspace*{-0.1in}
In the above code, the instantiation of $\rhoalloc$ and $\rho_1$ with
$\rgn_1$, and $\rho_2$ with $\rgn_0$ is allowed because (a) $\rgn_0$
and $\rgn_1$ are live during the instantiation, and (b). $\rgn_0
\outlives \rgn_1$ and $\rgn_1 \outlives \rgn_1$ (since outlives is
reflexive). Observe that the region type of \C{p} conveys
the fact that (a). it is allocated in region $\rgn_1$, and (b). it
holds references to objects allocated in region $\rgn_0$ and $\rgn_1$.
In contrast, if we choose to allocate the \C{snd} object also in
$\rgn_1$, then \C{p} would be contained in $\rgn_1$, and its region
type would be \C{Pair<\ObjZ,\ObjZ><$\rgn_1$,$\rgn_1$,$\rgn_1$>}, which
we abbreviate as \C{Pair<\ObjZ,\ObjZ>@$\rgn_1$}. {In general, we treat
$B\inang{\tbar}@\rgn$ as being equivalent to
$B\inang{\tbar}\inang{\rbar}$}. Region annotation on type $a$, where
$a$ is a type variable, assumes the form $a@\rgn$. If $a$ is
instantiated with \C{Pair<\ObjZ,\ObjZ>}, the result is the type of a
\C{Pair} object contained in  $\rgn$. The type $\unitZ$ is unboxed in
\FB, hence it has no region annotations.
% The region type of $\ObjZ$ class in \FB has no fields, hence its
% region type is of form $\ObjZ@\rgn$. 
%Class definitions in \FB extend FGJ's with region parameters
%($\rho$). Informally, region variables ($\rho$) are to region
%identifiers ($\rgn$) as type variables ($a$) are to core types ($T$);
%they can be instantiated with region identifiers, as needed. A class
%in \FB is necessarily parameterized over the allocation region of its
%objects. We use $\rho^a$ to denote this parameter, and often refer to
%it as \emph{class}'s allocation region, although it is really the
%allocation region of its objects. An Object in \FB can contain fields
%referring to objects allocated in regions ($\rhobar$) other than its
%own allocation region ($\rhoalloc$), provided that the former outlive
%the later (i.e., $\rhobar \outlives \rho$). In such cases, the
%definition of object's class needs to be parametric over allocation
%regions of its fields (i.e., their classes). Furthermore, the
%constraint that such regions must outlive the allocation region of the
%class needs to be made explicit in the definition. For instance, a
%\C{Pair} object allocated in a region ${\rhoalloc}$ can have its first
%component coming from ${\rho_1}$ and second from $\rho_2$, where all
%the $\rho$'s are distinct, and $\rho_1, \rho_2$ outlive
%$\rhoalloc$. T

@@

Furthermore, we prove the following theorem about \FB, which, in
conjunction with the type safety theorem, implies the safety of region
transfers across address spaces:
\begin{theorem}
\emph{(\textbf{Transfer Safety})}
\label{thm:fb-transfer-safety}
$\forall$ $v$, $\rhoenv$, $\rhoenv'$, $\rhomap$, $\rhomap'$, $\phicx$,
$\phicx'$, $\rgn$, $\rgn'$, such that $\rgn \in \rhoenv$, $\rgn' \in
\rhoenv'$, $\rgn_i \neq \toprgn$ and
$\rgn_i \notin dom(\rhomap') \cup \rhoenv'$, if $\hastyp{\emptyA,\rgn,
\cdot}{\C{new}\; \RgnZT{\rgn_i}(v)}{\RgnZT{\toprgn}}$, then
$\hastyp{(\rhomap'[\rgn_i \mapsto \CLOSED],\rhoenv',\cdot,\phicx')}
{\\\C{new}\; \RgnZT{\rgn_i}(v)}{\RgnZT{\toprgn}}$
\end{theorem}
The above theorem states that if a $\C{new}\; \RgnZT{\rgn_i}(v)$ value
is well-typed under one context, then it is also well-typed under
every other context, whose $\rhomap$ maps $\rgn_i$ to closed
($\CLOSED$) state.  Thus, a recipient of a transferable region only
needs to add a binding for the region to its $\rhomap$ in order to
preserve its type safety. 
% Notably, this result could not have been
% established if it were possible for a transferable region to contain
% references to objects outside the region.

>>>>

To construct objects of the \C{Pair} class, its type and region
parameters need to be instantiated with core types ($T$) and concrete
region identifiers\footnote{Region variables ($\rho$) and region
identifiers ($\rgn$) belong to the same syntactic class. Region
identifiers are to region variables, as types ($T$) are to type
variables ($a$)} ($\rgn$), respectively. For example:
\begin{codejava}
letregion $\rgn_0$ in
  let snd = new Object<$\rgn_0$>() in
  letregion $\rgn_1$ in
    let fst = new Object<$\rgn_1$>() in
    let p = new Pair<Object,Object><$\rgn_1$,$\rgn_1$,$\rgn_0$>
                  (fst,snd);
\end{codejava}
%\vspace*{-0.1in}
In the above code, the instantiation of $\rhoalloc$ and $\rho_1$ with
$\rgn_1$, and $\rho_2$ with $\rgn_0$ is allowed because (a) $\rgn_0$
and $\rgn_1$ are live during the instantiation, and (b). $\rgn_0
\outlives \rgn_1$ and $\rgn_1 \outlives \rgn_1$ (since outlives is
reflexive). Observe that the region type of \C{p} conveys
the fact that (a). it is allocated in region $\rgn_1$, and (b). it
holds references to objects allocated in region $\rgn_0$ and $\rgn_1$.
In contrast, if we choose to allocate the \C{snd} object also in
$\rgn_1$, then \C{p} would be contained in $\rgn_1$, and its region
type would be \C{Pair<\ObjZ,\ObjZ><$\rgn_1$,$\rgn_1$,$\rgn_1$>}, which
we abbreviate as \C{Pair<\ObjZ,\ObjZ>@$\rgn_1$}. In general, we treat
$B\inang{\tbar}@\rgn$ as being equivalent to
$B\inang{\tbar}\inang{\rbar}$. Region annotation on type $a$, where
$a$ is a type variable, assumes the form $a@\rgn$. If $a$ is
instantiated with \C{Pair<\ObjZ,\ObjZ>}, the result is the type of a
\C{Pair} object contained in  $\rgn$. Separate parameterization over
types and regions is needed in \FB to support region-polymorphic
higher-order functions (explained below) as class fields. Class fields
are not allowed to be type polymorphic, in keeping with C\#, which
constitutes the core of \name. For class fields that are functions,
regions are parameterized at the (class) field level whereas types are
parameterized at the class level.  This choice also simplifies type
inference in the presence of higher-order functions. The type $\unitZ$
is unboxed in \FB, hence it has no region annotations.

% The region type of $\ObjZ$ class in \FB has no fields, hence its
% region type is of form $\ObjZ@\rgn$. 
%Class definitions in \FB extend FGJ's with region parameters
%($\rho$). Informally, region variables ($\rho$) are to region
%identifiers ($\rgn$) as type variables ($a$) are to core types ($T$);
%they can be instantiated with region identifiers, as needed. A class
%in \FB is necessarily parameterized over the allocation region of its
%objects. We use $\rho^a$ to denote this parameter, and often refer to
%it as \emph{class}'s allocation region, although it is really the
%allocation region of its objects. An Object in \FB can contain fields
%referring to objects allocated in regions ($\rhobar$) other than its
%own allocation region ($\rhoalloc$), provided that the former outlive
%the later (i.e., $\rhobar \outlives \rho$). In such cases, the
%definition of object's class needs to be parametric over allocation
%regions of its fields (i.e., their classes). Furthermore, the
%constraint that such regions must outlive the allocation region of the
%class needs to be made explicit in the definition. For instance, a
%\C{Pair} object allocated in a region ${\rhoalloc}$ can have its first
%component coming from ${\rho_1}$ and second from $\rho_2$, where all
%the $\rho$'s are distinct, and $\rho_1, \rho_2$ outlive
%$\rhoalloc$. T

@@

Furthermore, we prove the following theorem about \FB, which, in
conjunction with the type safety theorem, implies the safety of region
transfers across address spaces:
\begin{theorem}
\emph{(\textbf{Transfer Safety})}
\label{thm:fb-type-safety}
$\forall$ $v$, $\rhoenv$, $\rhoenv'$, $\rhomap$, $\rhomap'$, $\phicx$,
$\phicx'$, $\rgn$, $\rgn'$, such that $\rgn \in \rhoenv$, $\rgn' \in
\rhoenv'$, $\rgn_i \neq \toprgn$ and
$\rgn_i \notin dom(\rhomap') \cup \rhoenv'$, if $\hastyp{\emptyA,\rgn,
\cdot}{\C{new}\; \RgnZT{\rgn_i}(v)}{\RgnZT{\toprgn}}$, then
$\hastyp{(\rhomap'[\rgn_i \mapsto \CLOSED],\rhoenv',\cdot,\phicx')}
{\\\C{new}\; \RgnZT{\rgn_i}(v)}{\RgnZT{\toprgn}}$
\end{theorem}
The above theorem states that if a $\C{new}\; \RgnZT{\rgn_i}(v)$ value
is well-typed under one context, then it is also well-typed under
every other context, whose $\rhomap$ maps $\rgn_i$ to closed
($\CLOSED$) state.  Thus, a recipient of a transferable region only
needs to add a binding for the region to its $\rhomap$ in order to
preserve its type safety. 
% Notably, this result could not have been
% established if it were possible for a transferable region to contain
% references to objects outside the region.
